---
title: "Sensitivity Analysis for Conservation Decision-Making"
subtitle: "Which parameter drives Ne/N most — and what can a manager actually change?"
author: "Raymond L. Tremblay"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    mathjax: default
    theme: flatly
    highlight: tango
vignette: >
  %\VignetteIndexEntry{Sensitivity Analysis for Conservation Decision-Making}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5,
  warning   = FALSE,
  message   = FALSE
)
```

```{r load-nestage, include=FALSE}
library(NeStage)
```

```{r packages, include=FALSE}
library(ggplot2)
library(dplyr)
library(purrr)
library(tidyr)
library(knitr)
```

---

# Introduction

## From a ratio to a decision

The previous vignette (*NeStage Functions: A User Guide*) established that all
three *Lepanthes* orchid species in Puerto Rico have Ne/N ratios far below the
short-term inbreeding threshold of Ne ≥ 50, even at census sizes of N = 40
— a large population for these species. Knowing that Ne/N = 0.18 for a
particular population tells you the situation is serious. It does not tell you
**what to do about it**.

Sensitivity analysis answers the management question: **which parameter, if
changed, would increase Ne/N the most?** This matters because managers have
limited resources and can realistically influence only certain aspects of a
population's biology:

| What a manager can influence | Relevant parameter |
|---|---|
| Reducing harvest or disturbance of reproductive adults | Stage frequency D — more adults |
| Translocating individuals between populations | Census N, D |
| Ex-situ propagation with controlled pollination | Reproductive variance Vk/k̄ |
| Removing dominant clonal genets | Clonal fraction d, Vc/c̄ |
| Improving habitat to extend plant lifespan | Generation time L |

## The four sensitivity functions

`NeStage` provides four sensitivity functions, each sweeping one parameter
while holding all others fixed:

| Function | Parameter swept | Applies to |
|---|---|---|
| `Ne_sensitivity_Vk()` | Sexual reproductive variance Vk/k̄ | Sexual, mixed |
| `Ne_sensitivity_Vc()` | Clonal reproductive variance Vc/c̄ | Mixed only |
| `Ne_sensitivity_d()` | Clonal fraction d | Mixed only |
| `Ne_sensitivity_L()` | Generation time L | All three models |

Each function returns a data frame, a ggplot, and an `Ne_sensitivity` object
with elasticity statistics from `print()`.

## Elasticity: a standardised sensitivity measure

Raw sensitivity (how much does Ne/N change per unit change in a parameter) is
hard to compare across parameters measured in different units. **Elasticity**
standardises by expressing the response as a proportional change:

$$
E = \frac{\partial (N_e/N)}{\partial \theta} \times \frac{\theta_\text{ref}}{(N_e/N)_\text{ref}}
$$

An elasticity of −0.4 means a 10% increase in $\theta$ produces a 4% decrease
in Ne/N. Elasticities are dimensionless and directly comparable across
parameters and populations.

---

# Study populations

We use three representative *Lepanthes* populations from the previous vignette —
one from each species — chosen to span the range of Ne/N values observed.

```{r setup-populations}
# ---- L. rupestris population 1 (Ne/N = 0.572) ----
MatU_Lrup1 <- matrix(c(
  0,     0,     0,     0,     0,     0,
  0.738, 0.738, 0,     0,     0,     0,
  0,     0,     0.515, 0,     0.076, 0.013,
  0,     0.038, 0,     0.777, 0,     0,
  0,     0.002, 0.368, 0.011, 0.730, 0.171,
  0,     0,     0.037, 0,     0.169, 0.790
), nrow = 6, byrow = TRUE,
dimnames = list(paste0("s", 1:6), paste0("s", 1:6)))

F_Lrup1 <- c(0, 0, 0, 0, 0.120, 0.414)   # monthly fecundity
D_Lrup1 <- c(22, 22, 36, 36, 48.5, 48.5)
D_Lrup1 <- D_Lrup1 / sum(D_Lrup1)

# ---- L. rubripetala population 1 (Ne/N = 0.930) ----
MatU_Lrub1 <- matrix(c(
  0,     0,     0,     0,     0,
  0.667, 0.667, 0,     0,     0,
  0,     0,     0.825, 0,     0.227,
  0,     0.037, 0,     0.812, 0,
  0,     0,     0.137, 0.010, 0.739
), nrow = 5, byrow = TRUE,
dimnames = list(paste0("s", 1:5), paste0("s", 1:5)))

F_Lrub1 <- c(0, 0, 0, 0, 0.043)
N_Lrub1 <- c(9/2, 9/2, 44/2, 44/2, 101)
D_Lrub1 <- N_Lrub1 / sum(N_Lrub1)

# ---- L. rubripetala population 5 (Ne/N = 0.250, lowest in Lrub) ----
MatU_Lrub5 <- matrix(c(
  0,     0,     0,     0,     0,
  0.667, 0.667, 0,     0,     0,
  0,     0,     0.726, 0,     0.250,
  0,     0.037, 0,     0.812, 0,
  0,     0,     0.237, 0.010, 0.739
), nrow = 5, byrow = TRUE,
dimnames = list(paste0("s", 1:5), paste0("s", 1:5)))

F_Lrub5 <- c(0, 0, 0, 0, 0.067)
N_Lrub5 <- c(52/2, 52/2, 4/2, 4/2, 46)
D_Lrub5 <- N_Lrub5 / sum(N_Lrub5)

# Baseline Ne estimates for reference
Ne_base_Lrup1 <- Ne_sexual_Y2000(
  T_mat = MatU_Lrup1, F_vec = F_Lrup1, D = D_Lrup1,
  Ne_target = 50, census_N = 40, population = "L. rupestris pop 1"
)
Ne_base_Lrub1 <- Ne_sexual_Y2000(
  T_mat = MatU_Lrub1, F_vec = F_Lrub1, D = D_Lrub1,
  Ne_target = 50, census_N = 40, population = "L. rubripetala pop 1"
)
Ne_base_Lrub5 <- Ne_sexual_Y2000(
  T_mat = MatU_Lrub5, F_vec = F_Lrub5, D = D_Lrub5,
  Ne_target = 50, census_N = 40, population = "L. rubripetala pop 5"
)

# Baseline summary
data.frame(
  Population = c("L. rupestris pop 1", "L. rubripetala pop 1",
                 "L. rubripetala pop 5"),
  Stages     = c(6, 5, 5),
  NeN        = round(c(Ne_base_Lrup1$NeN, Ne_base_Lrub1$NeN,
                        Ne_base_Lrub5$NeN), 3),
  Ne_at_40   = c(Ne_base_Lrup1$Ne_at_census, Ne_base_Lrub1$Ne_at_census,
                  Ne_base_Lrub5$Ne_at_census),
  Min_N_50   = c(Ne_base_Lrup1$Min_N, Ne_base_Lrub1$Min_N,
                  Ne_base_Lrub5$Min_N)
) |>
  knitr::kable(
    col.names = c("Population", "Stages", "Ne/N",
                  "Ne at N=40", "Min N (Ne≥50)"),
    caption   = "Baseline Ne/N estimates for the three study populations
                 (monthly matrices, Ne_target = 50, census_N = 40)."
  )
```

---

# Sensitivity to sexual reproductive variance: `Ne_sensitivity_Vk()`

## What is Vk/k̄ and why does it matter?

The variance-to-mean ratio of sexual reproductive output Vk/k̄ measures how
equally seeds are distributed among reproductive adults. Under Poisson
distribution (Vk/k̄ = 1), all adults contribute equally to the seed pool.
Values above 1 mean some individuals dominate — a few adults produce most of
the seeds, so the effective number of parents is smaller than the census count.

In *Lepanthes*, this is biologically plausible because:

- Pollination by fungus gnats is patchy — some plants are visited far more
  frequently than others
- Microhabitat variation means some individuals are far more fecund than others
- In small populations, one or two dominant individuals can produce most seeds
  in a given year

The Poisson default (Vk/k̄ = 1) is almost certainly an underestimate of the
true variance for most natural populations.

## Running `Ne_sensitivity_Vk()` for the reproductive stage

For *L. rupestris* (6 stages), stage 5 (single-shoot reproductive adults)
is the primary sexual reproductive stage. We sweep Vk/k̄ from 0.5 to 8.

```{r vk-sensitivity-lrup1, fig.cap="Sensitivity of Ne/N to sexual reproductive variance in stage 5 of *L. rupestris* pop 1. Dashed vertical line = Poisson default."}
sens_Vk_Lrup1 <- Ne_sensitivity_Vk(
  model_fn    = Ne_sexual_Y2000,
  T_mat       = MatU_Lrup1,
  F_vec       = F_Lrup1,
  D           = D_Lrup1,
  stage_index = 5,              # stage 5: single-shoot reproductive adults
  Vk_range    = seq(0.5, 8, by = 0.5),
  Ne_target   = 50,
  population  = "L. rupestris pop 1"
)

print(sens_Vk_Lrup1)
```

## Interpretation

The elasticity printed above tells us how strongly Ne/N responds to Vk/k̄.
A negative local elasticity is expected — higher reproductive variance always
reduces Ne/N. The conservation ratio (worst-case Ne/N relative to Poisson
baseline) tells a manager: if reproductive inequality is at the high end of
the plausible range, how many more individuals are needed?

```{r vk-both-stages, fig.cap="Sensitivity of Ne/N to sexual reproductive variance: stage 5 (left) vs stage 6 (right) in *L. rupestris* pop 1.", fig.width=10}
# Also sweep stage 6 (two-or-more-shoot adults) for comparison
sens_Vk_Lrup1_s6 <- Ne_sensitivity_Vk(
  model_fn    = Ne_sexual_Y2000,
  T_mat       = MatU_Lrup1,
  F_vec       = F_Lrup1,
  D           = D_Lrup1,
  stage_index = 6,
  Vk_range    = seq(0.5, 8, by = 0.5),
  Ne_target   = 50,
  population  = "L. rupestris pop 1 — stage 6"
)

# Combined data for comparison plot
df_vk_compare <- bind_rows(
  sens_Vk_Lrup1$data    %>% mutate(Stage = "Stage 5 (1-shoot repro)"),
  sens_Vk_Lrup1_s6$data %>% mutate(Stage = "Stage 6 (2+-shoot repro)")
)

ggplot(df_vk_compare, aes(x = Vk_over_k, y = NeN, color = Stage)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("#2166ac", "#d6604d")) +
  labs(
    title    = expression(paste("Ne/N vs. sexual reproductive variance  ", V[k]/bar(k))),
    subtitle = "L. rupestris pop 1 — comparing focal stage",
    x        = expression(V[k]/bar(k)),
    y        = expression(N[e]/N),
    color    = NULL,
    caption  = "Dashed line = Poisson default (Vk/k̄ = 1). Monthly matrices."
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom",
    panel.grid.major.y = element_line(color = "grey92", linewidth = 0.4)
  )
```

Stage 6 (multi-shoot plants) has a larger effect because its fecundity
(F = 0.414 per month) is more than three times that of stage 5 (F = 0.120).
When high-fecundity adults have unequal reproductive output, the genetic
consequences are proportionally larger.

**Management implication:** Strategies that equalise seed production — for
example, hand-pollination programmes that ensure all reproductive adults
contribute pollen — would have the greatest effect in stage 6 individuals.

---

# Sensitivity to generation time: `Ne_sensitivity_L()`

## Why L matters so much

Generation time L appears in the denominator of the Ne/N formula (Eq. 6 of
Yonezawa et al. 2000): Ne/N = 2/(V·L). A longer generation time means that
drift accumulates over more years per generation, reducing Ne/N. Conversely,
**anything that shortens the generation time** — such as habitat degradation
that reduces adult survival — also reduces Ne/N.

Generation time is also the most **uncertain** parameter in the model. Field
studies of *Lepanthes* run for 16–34 months, which is far shorter than the
generation time of these long-lived orchids. The L value we compute from
monthly matrices may differ from the true population L if the observed years
were not demographically representative.

## Sweeping L across a biologically plausible range

```{r L-sensitivity, fig.cap="Sensitivity of Ne/N to generation time L across all three study populations.", fig.width=10, fig.height=5}
L_range_use <- seq(5, 80, by = 5)

# Run for all three populations
make_L_sens <- function(MatU, F_vec, D, Ne_target, pop_label) {
  Ne_sensitivity_L(
    model_fn   = Ne_sexual_Y2000,
    T_mat      = MatU,
    F_vec      = F_vec,
    D          = D,
    L_range    = L_range_use,
    Ne_target  = Ne_target,
    population = pop_label
  )
}

sens_L_Lrup1 <- make_L_sens(MatU_Lrup1, F_Lrup1, D_Lrup1, 50,
                              "L. rupestris pop 1")
sens_L_Lrub1 <- make_L_sens(MatU_Lrub1, F_Lrub1, D_Lrub1, 50,
                              "L. rubripetala pop 1")
sens_L_Lrub5 <- make_L_sens(MatU_Lrub5, F_Lrub5, D_Lrub5, 50,
                              "L. rubripetala pop 5")

# Observed L values from monthly matrices
L_obs <- c(Ne_base_Lrup1$L, Ne_base_Lrub1$L, Ne_base_Lrub5$L)

df_L <- bind_rows(
  sens_L_Lrup1$data %>% mutate(Population = "L. rupestris pop 1",
                                 L_obs = L_obs[1]),
  sens_L_Lrub1$data %>% mutate(Population = "L. rubripetala pop 1",
                                 L_obs = L_obs[2]),
  sens_L_Lrub5$data %>% mutate(Population = "L. rubripetala pop 5",
                                 L_obs = L_obs[3])
)

ggplot(df_L, aes(x = L, y = NeN, color = Population)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  geom_vline(aes(xintercept = L_obs, color = Population),
             linetype = "dashed", linewidth = 0.7, alpha = 0.7) +
  scale_color_manual(values = c("#2166ac", "#1b7837", "#d6604d")) +
  labs(
    title    = expression(paste("Sensitivity of  ", N[e]/N, "  to generation time  ", L)),
    subtitle = "Dashed vertical lines = observed L from monthly matrices",
    x        = "Generation time L (months)",
    y        = expression(N[e]/N),
    color    = NULL,
    caption  = "All populations: Ne_sexual_Y2000, Poisson defaults, Ne_target = 50."
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold"),
    legend.position = "bottom",
    legend.text     = element_text(face = "italic"),
    panel.grid.major.y = element_line(color = "grey92", linewidth = 0.4)
  )
```

```{r L-print}
# Print elasticity summary for the most sensitive population
print(sens_L_Lrub5)
```

## Key insight: Ne/N is hyperbolic in L

Because Ne/N = 2/(V·L), the relationship is hyperbolic — Ne/N declines
steeply at short L and flattens at long L. This means:

- Populations with **short** observed L (such as *L. rubripetala* pop 5 with
  L ≈ 20 months) are in the steep part of the curve — a small error in L
  estimation translates into a large error in Ne/N.
- Populations with **longer** L are more stable in their Ne/N estimate.

**Management implication:** For populations where L is highly uncertain, it is
worth investing in longer-term demographic monitoring (ideally ≥ 5 years) to
pin down L more precisely before interpreting Ne/N.

---

# Sensitivity to clonal fraction: `Ne_sensitivity_d()`

## When does clonal fraction matter?

A fundamental and counter-intuitive result of Yonezawa et al. (2000) is that
**under Poisson defaults (Vc/c̄ = Vk/k̄ = 1), the clonal fraction d has no
effect on Ne/N**. This is because when all individuals reproduce equally
(whether sexually or clonally), the mode of reproduction does not matter for
genetic drift.

Clonal fraction only reduces Ne/N when **clonal output is more variable than
sexual output** — when some genets dominate clonal reproduction, reducing the
number of genetically distinct lineages. We demonstrate both cases here using
a hypothetical mixed-reproduction *Lepanthes* population.

```{r d-sensitivity, fig.cap="Effect of clonal fraction on Ne/N: flat under Poisson variance (left), declining under super-Poisson clonal variance (right).", fig.width=10}
# Use L. rubripetala pop 1 matrix structure with a mixed reproduction scenario
# Add a clonal component: imagining a hypothetical scenario where stage 5
# (reproductive adults) can reproduce clonally via offshoots

# Poisson case — d has no effect
sens_d_poisson <- Ne_sensitivity_d(
  T_mat       = MatU_Lrub1,
  F_vec       = F_Lrub1,
  D           = D_Lrub1,
  d_fixed     = c(0, 0, 0, 0, 0.5),   # base: stage 5 is 50% clonal
  stage_index = 5,
  d_range     = seq(0, 1, by = 0.1),
  Vc_over_c   = rep(1, 5),             # Poisson clonal variance
  Ne_target   = 50,
  population  = "L. rubripetala pop 1 (Poisson)"
)

# Super-Poisson clonal variance — d starts to matter
sens_d_superpoisson <- Ne_sensitivity_d(
  T_mat       = MatU_Lrub1,
  F_vec       = F_Lrub1,
  D           = D_Lrub1,
  d_fixed     = c(0, 0, 0, 0, 0.5),
  stage_index = 5,
  d_range     = seq(0, 1, by = 0.1),
  Vc_over_c   = c(1, 1, 1, 1, 4),     # stage 5 clonal output highly variable
  Ne_target   = 50,
  population  = "L. rubripetala pop 1 (Vc/c̄ = 4)"
)

df_d <- bind_rows(
  sens_d_poisson$data      %>% mutate(Scenario = "Poisson (Vc/c̄ = 1)"),
  sens_d_superpoisson$data %>% mutate(Scenario = "Super-Poisson (Vc/c̄ = 4)")
)

ggplot(df_d, aes(x = d_focal, y = NeN, color = Scenario)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_vline(xintercept = 0.5, linetype = "dashed",
             color = "grey50", linewidth = 0.7) +
  annotate("text", x = 0.51, y = max(df_d$NeN) * 0.98,
           label = "Base d = 0.5", hjust = 0, size = 3.2, color = "grey40") +
  scale_color_manual(values = c("#1b7837", "#d6604d")) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.2)) +
  labs(
    title    = expression(paste("Ne/N vs. clonal fraction  ", d[i], "  in stage 5")),
    subtitle = "L. rubripetala pop 1 — hypothetical mixed reproduction scenario",
    x        = expression(paste("Clonal fraction  ", d[i],
                                "  (0 = fully sexual,  1 = fully clonal)")),
    y        = expression(N[e]/N),
    color    = NULL,
    caption  = "Ne_mixed_Y2000. Flat curve confirms Poisson result of Yonezawa et al. (2000)."
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold"),
    legend.position = "bottom",
    panel.grid.major.y = element_line(color = "grey92", linewidth = 0.4)
  )
```

```{r d-print-superpoisson}
print(sens_d_superpoisson)
```

**Management implication:** Simply removing clonal propagules (to reduce d)
has no genetic benefit under Poisson variance assumptions. The conservation
benefit of managing clonal reproduction comes only when dominant genets can
be identified and their disproportionate contribution to the clonal pool
reduced — which requires molecular markers to identify genets in the field.

---

# Cross-parameter comparison: which lever matters most?

We now compare the local elasticities across all parameters and all three
populations to produce a prioritisation framework.

```{r elasticity-comparison}
# Extract local elasticity from each sensitivity object
# Local elasticity approximated by central finite difference at Vk/k_bar = 1
# E = (dNeN/dtheta) * (theta_ref / NeN_ref)

local_elasticity <- function(sens_obj, param_col, ref_val) {
  df  <- sens_obj$data
  idx <- which.min(abs(df[[param_col]] - ref_val))
  # Use neighbouring points for central difference
  if (idx == 1)        idx <- 2
  if (idx == nrow(df)) idx <- nrow(df) - 1
  dNeN   <- df$NeN[idx + 1] - df$NeN[idx - 1]
  dtheta <- df[[param_col]][idx + 1] - df[[param_col]][idx - 1]
  NeN_ref <- df$NeN[idx]
  (dNeN / dtheta) * (ref_val / NeN_ref)
}

# L. rupestris pop 1: sweep Vk stages 5 and 6, and L
E_Vk5_Lrup1 <- local_elasticity(sens_Vk_Lrup1,    "Vk_over_k", 1)
E_Vk6_Lrup1 <- local_elasticity(sens_Vk_Lrup1_s6, "Vk_over_k", 1)
E_L_Lrup1   <- local_elasticity(sens_L_Lrup1,      "L",
                                  Ne_base_Lrup1$L)

# L. rubripetala pop 1
sens_Vk_Lrub1 <- Ne_sensitivity_Vk(
  model_fn    = Ne_sexual_Y2000,
  T_mat       = MatU_Lrub1,
  F_vec       = F_Lrub1,
  D           = D_Lrub1,
  stage_index = 5,
  Vk_range    = seq(0.5, 8, by = 0.5),
  Ne_target   = 50,
  population  = "L. rubripetala pop 1"
)
E_Vk5_Lrub1 <- local_elasticity(sens_Vk_Lrub1, "Vk_over_k", 1)
E_L_Lrub1   <- local_elasticity(sens_L_Lrub1,   "L", Ne_base_Lrub1$L)

# L. rubripetala pop 5
sens_Vk_Lrub5 <- Ne_sensitivity_Vk(
  model_fn    = Ne_sexual_Y2000,
  T_mat       = MatU_Lrub5,
  F_vec       = F_Lrub5,
  D           = D_Lrub5,
  stage_index = 5,
  Vk_range    = seq(0.5, 8, by = 0.5),
  Ne_target   = 50,
  population  = "L. rubripetala pop 5"
)
E_Vk5_Lrub5 <- local_elasticity(sens_Vk_Lrub5, "Vk_over_k", 1)
E_L_Lrub5   <- local_elasticity(sens_L_Lrub5,   "L", Ne_base_Lrub5$L)

# Assemble table
elast_tbl <- data.frame(
  Population = c("*L. rupestris* pop 1",
                 "*L. rubripetala* pop 1",
                 "*L. rubripetala* pop 5"),
  Base_NeN   = round(c(Ne_base_Lrup1$NeN,
                         Ne_base_Lrub1$NeN,
                         Ne_base_Lrub5$NeN), 3),
  E_Vk_repro = round(c(E_Vk5_Lrup1, E_Vk5_Lrub1, E_Vk5_Lrub5), 3),
  E_L        = round(c(E_L_Lrup1,   E_L_Lrub1,   E_L_Lrub5),   3)
)

knitr::kable(elast_tbl,
  col.names = c("Population", "Base Ne/N",
                "E(Vk/k̄) repro stage", "E(L)"),
  escape  = FALSE,
  caption = "Local elasticities at the Poisson reference point (Vk/k̄ = 1)
             and at the observed L. E(Vk/k̄) < 0: increasing reproductive
             variance reduces Ne/N. E(L) < 0: longer generation time
             reduces Ne/N. Larger absolute values indicate higher sensitivity."
)
```

```{r elasticity-plot, fig.cap="Absolute local elasticity of Ne/N with respect to two parameters across three populations. Larger bars indicate parameters where management intervention would have the greatest proportional effect on Ne/N.", fig.height=4}
elast_long <- elast_tbl %>%
  select(Population, E_Vk_repro, E_L) %>%
  pivot_longer(cols = c(E_Vk_repro, E_L),
               names_to = "Parameter",
               values_to = "Elasticity") %>%
  mutate(
    Abs_E     = abs(Elasticity),
    Parameter = recode(Parameter,
                       E_Vk_repro = "Sexual repro variance\n(Vk/k̄, repro stage)",
                       E_L        = "Generation time (L)")
  )

ggplot(elast_long, aes(x = Population, y = Abs_E, fill = Parameter)) +
  geom_col(position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("#2166ac", "#d6604d")) +
  scale_x_discrete(labels = function(x) gsub("\\*", "", x)) +
  labs(
    title    = "|Elasticity| of Ne/N: which parameter drives genetic drift most?",
    subtitle = "Higher = Ne/N more sensitive to this parameter",
    x        = NULL,
    y        = "|Local elasticity|",
    fill     = "Parameter",
    caption  = "Local elasticity at Poisson reference (Vk/k̄ = 1) and observed L.
                All populations: Ne_sexual_Y2000, monthly matrices."
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold"),
    axis.text.x     = element_text(face = "italic", size = 9),
    legend.position = "right",
    panel.grid.major.y = element_line(color = "grey92", linewidth = 0.4)
  )
```

---

# Conservation prioritisation framework

The sensitivity analysis above leads to the following prioritisation framework
for *Lepanthes* conservation, ordered by the expected genetic return on
management investment.

## Priority 1 — Reduce uncertainty in L

Generation time L is the most influential parameter in all three populations
and the most poorly estimated. Monthly matrices from 16–34 months of data
capture only 1–3 generation cycles for these long-lived orchids. A dedicated
mark-recapture programme of at least 5 years per population would substantially
reduce uncertainty in L and allow more confident Ne/N estimates.

## Priority 2 — Equalise reproductive output in high-fecundity stages

The elasticity of Ne/N to Vk/k̄ is consistently negative and non-trivial.
In *L. rupestris*, where stage 6 (multi-shoot adults) has the highest
fecundity (F = 0.414/month), ensuring that these adults all contribute to
the seed pool — via controlled hand-pollination in ex situ programmes, or
by removing physical barriers to pollinator movement — would increase Ne/N.

## Priority 3 — Augment census size in the lowest-Ne/N populations

The minimum census size to achieve Ne ≥ 50 ranges from 54 (*L. rubripetala*
pop 1) to 314 (*L. rupestris* pop 3). Translocation of individuals from
source populations to these sinks would not only increase N directly, but
would also restore genetic diversity lost to drift.

```{r final-comparison, fig.cap="Ne/N as a function of Vk/k̄ in the reproductive stage for all three study populations. Populations with higher baseline Ne/N are also less sensitive to reproductive variance.", fig.height=5}
df_final <- bind_rows(
  sens_Vk_Lrup1$data %>% mutate(Population = "*L. rupestris* pop 1"),
  sens_Vk_Lrub1$data %>% mutate(Population = "*L. rubripetala* pop 1"),
  sens_Vk_Lrub5$data %>% mutate(Population = "*L. rubripetala* pop 5")
)

ggplot(df_final, aes(x = Vk_over_k, y = NeN, color = Population)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_vline(xintercept = 1, linetype = "dashed",
             color = "grey50", linewidth = 0.7) +
  geom_hline(yintercept = 50/40, linetype = "dotted",
             color = "grey30", linewidth = 0.6) +
  annotate("text", x = 7.5, y = 50/40 + 0.03,
           label = "Ne=50 at N=40", hjust = 1, size = 3, color = "grey30") +
  scale_color_manual(
    values = c("#2166ac", "#1b7837", "#d6604d"),
    labels = c(
      expression(italic("L. rubripetala")~"pop 1"),
      expression(italic("L. rubripetala")~"pop 5"),
      expression(italic("L. rupestris")~"pop 1")
    )
  ) +
  labs(
    title    = expression(paste("Ne/N vs. sexual reproductive variance  ",
                                V[k]/bar(k), "  — three populations")),
    subtitle = "Dotted horizontal line = Ne = 50 threshold at N = 40",
    x        = expression(V[k]/bar(k)~"(reproductive stage)"),
    y        = expression(N[e]/N),
    color    = NULL,
    caption  = "Ne_sexual_Y2000, monthly matrices, Ne_target = 50, census_N = 40."
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold"),
    legend.position = "bottom",
    panel.grid.major.y = element_line(color = "grey92", linewidth = 0.4)
  )
```

---

# Session information

```{r session-info}
sessionInfo()
```

---

# References

- Caswell H. (2001). *Matrix Population Models*, 2nd ed. Sinauer Associates.
- Franklin I.R. (1980). Evolutionary change in small populations. In: Soulé &
  Wilcox, eds. *Conservation Biology*. Sinauer. pp. 135–149.
- Frankham R. (1995). Effective population size/adult population size ratios
  in wildlife: a review. *Genetical Research* **66**: 95–107.
- Lande R. (1995). Mutation and conservation. *Conservation Biology* **9**:
  782–791.
- Morris W.F. & Doak D.F. (2002). *Quantitative Conservation Biology*.
  Sinauer Associates. Ch. 9.
- Tremblay R.L. & Ackerman J.D. (2001). Gene flow and effective population
  size in *Lepanthes* (Orchidaceae): a case for genetic drift.
  *Biological Journal of the Linnean Society* **72**: 47–62.
- Wright S. (1931). Evolution in Mendelian populations. *Genetics* **16**:
  97–159.
- Yonezawa K., Ishida T. & Iwata H. (2000). Formulation and estimation of
  the effective size of stage-structured populations. *Evolution* **54**:
  2007–2013.
