---
title: "NeStage: Reproducing Yonezawa (2000) Table 4 and Using the Exact Backend"
author: "Raymond L. Tremblay"
output: html_vignette
vignette: >
  %\VignetteIndexEntry{NeStage: Reproducing Yonezawa (2000) Table 4 and Using the Exact Backend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)

```

## Overview

This vignette demonstrates how to compute generation time (L), Ny/N, and Ne/N using the **exact Yonezawa (2000)** definitions as implemented in NeStage’s `"paper_v1"` backend. We:

-   Reproduce **Table 4** results for the **Mizuyajiri (Miz)** and **Nanryu (Nan)** populations using **observed** and **expected** stage fractions.

-   Show a simple **sex-only** example (illustrative).

-   Provide guidance for running the exact backend on **your own U, F, D** inputs (with optional L computation).

> Important: To reproduce Yonezawa Table 4 exactly, we use `meta$estimator = "paper_v1"` and the paper’s stage fractions (D) and L values. For new datasets, you can compute L internally by the paper’s age-by-stage definition when `L` is not provided.

## Loading NeStage

If you have installed the package, simply:

```{r}

```

For development (sourcing from `R/` files), use:

```{r}

```

## Reproducing Yonezawa (2000) Table 4

We use helper inputs for Miz/Nan (U, F, D_obs, D_exp, L) and call `nestage_exact()` with the exact backend.

```{r}

```

You should see values that match Table 4 (rounded):

-   

-   Miz: L = **13.399**, Ny/N = **2.932 (2.976)**, Ne/N = **0.219 (0.222)**

-   

-   Nan: L = **8.353**, Ny/N = **2.428 (2.446)**, Ne/N = **0.291 (0.293)**

-   

> Note on rounding: Formatting in the original table may reflect rounding at the third decimal. Small differences (±0.001–0.002) are expected depending on rounding conventions.

### Quick point-only plot (no bars)

We can visualize Ny/N and Ne/N for the four results (Miz/Nan and their parenthetical values) using **points** (not bars). If `ggplot2` is available we use it; otherwise base plots.

```{r}

```

## Sex-only example (illustrative)

This shows how to run a **sex-only** variant by scaling fecundity for a given stage while keeping the exact backend. Here we reduce stage-3 reproduction to 10% of the original for Miz, and we keep the observed stage fractions D_obs for reporting.

> Caution: Under extreme down-scaling of fecundity, `1 - overline(u^2)` can become very small, and `L` may increase, which can yield large Ny/N and Ne/N. These are mathematically consistent with the formulation but should be interpreted with care.

## Using your own U, F, D (and L)

For a general dataset:

-   

-   Provide `U` (square transition matrix) and `F` (length k fecundity vector).

-   

-   Provide stage fractions `D` (length k, non-negative, sum to 1) for reporting weights.

-   

-   Optionally provide `L`. If omitted, NeStage will compute L using the paper’s age-by-stage definition `L(U, F)`.

-   

-   Force the exact backend via `meta$estimator = "paper_v1"`.

### Stage-order alignment and assembly notes

1.  

2.  **Stage order** must be consistent across `U`, `F`, and `D`.

3.  

4.  **Lambda reporting** uses `A_full <- U; A_full[1, ] <- F` (row 1 is fecundity only).

5.  

6.  If you only have a full projection matrix `A` and no split into `U` and `F`, you can heuristically set `F <- A[1, ]` and `U <- A; U[1, ] <- 0`, but exact replication of Table 4 requires the correct paper `U` and `F`.

7.  

## Session info
