---
title: "Ne Across Plant Species: Using COMPADRE with NeStage"
author: "Raymond L. Tremblay"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_depth: 3
    number_sections: true
    mathjax: default
vignette: >
  %\VignetteIndexEntry{Ne Across Plant Species: Using COMPADRE with NeStage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = FALSE,       # set to TRUE only when COMPADRE is reachable
  message = TRUE,
  warning = FALSE,
  comment = "#>"
)
```

---

# Overview

This vignette shows how to compute the **variance effective population size
($N_e$)** for a large set of sexually reproducing plant species by connecting
NeStage directly to the [COMPADRE Plant Matrix Database](https://compadre-db.org).

Two wrapper functions handle everything:

| Function | What it does |
|---|---|
| `Ne_compadre_sexual_filter()` | Downloads COMPADRE and returns one clean matrix per species |
| `Ne_compadre_sexual()` | Runs `Ne_sexual_Y2000()` on each species and returns a tidy table |

**What you need:**

- An internet connection (to fetch COMPADRE)
- R packages: `Rcompadre`, `NeStage`, `ggplot2`, `dplyr`
- The file `Ne_compadre_sexual.R` sourced into your session

**What you get:**

- $N_e/N$ and minimum census size $N_{min}$ for every species that passes
  the quality filters
- Results grouped by plant family for comparative analyses
- Four ready-to-use `ggplot2` figures

---

# Background: why COMPADRE?

The COMPADRE Plant Matrix Database (Salguero-Gómez et al. 2015, 2016)
archives matrix population models (MPMs) from the published literature.
Each MPM is decomposed into three sub-matrices:

$$A = U + F + C$$

where $U$ is the survival/transition matrix (= `T_mat` in NeStage), $F$ is the
sexual fecundity matrix, and $C$ is the clonal reproduction matrix.

This decomposition maps directly onto what NeStage needs:

| NeStage input | COMPADRE source | How obtained |
|---|---|---|
| `T_mat` | $U$ matrix | `matU(db)[[i]]` |
| `F_vec` | Row 1 of $F$ matrix | `matF(db)[[i]][1, ]` |
| `D` | Not stored — computed | Stable stage distribution of $A = U + F$ |
| `L` | Not stored — computed | Internally by `Ne_sexual_Y2000()` |

The stable stage distribution $D$ (dominant right eigenvector of $A$,
normalised to sum to 1) is the theoretical equilibrium stage structure.
It is the most defensible default when field census data on stage frequencies
are unavailable, and is computed internally by `Ne_compadre_sexual()`.

---

# Setup

## Install packages

```{r install, eval=FALSE}
# Run once
install.packages("Rcompadre")
install.packages("ggplot2")
install.packages("dplyr")

# NeStage from GitHub
install.packages("remotes")
remotes::install_github("RaymondLTremblay/NeStage")
```

## Load packages and source the wrapper functions

```{r load}
library(Rcompadre)
library(NeStage)
library(ggplot2)
library(dplyr)

# Source the COMPADRE–NeStage bridge functions.
# Adjust the path to wherever you saved Ne_compadre_sexual.R
source("Ne_compadre_sexual.R")
```

---

# Step 1: Filter COMPADRE

`Ne_compadre_sexual_filter()` downloads the full COMPADRE database and
applies 12 sequential filters. With `verbose = TRUE` (the default) it prints
the number of matrices remaining after each step.

```{r filter}
db_sexual <- Ne_compadre_sexual_filter(verbose = TRUE)
```

The filter chain is:

1. `MatrixComposite == "Mean"` — mean matrices only (not individual annual matrices)
2. `MatrixTreatment == "Unmanipulated"` — no experimental treatment applied
3. `MatrixCaptivity == "W"` — wild populations only
4. `AnnualPeriodicity == 1` — annual projection interval (NeStage assumes 1-year steps)
5. `MatrixSplit == "Divided"` — $U$, $F$, $C$ sub-matrices are separately available
6. `MatrixFec == "Yes"` — fecundity was explicitly measured
7. `Kingdom == "Plantae"` — plants only
8. `check_NA_U == FALSE & check_NA_F == FALSE & check_NA_C == FALSE` — no missing values
9. `check_zero_U == FALSE` — at least one non-zero survival column
10. `check_zero_F == FALSE` — $F$ matrix has non-zero entries
11. `all(matC == 0)` — no clonal reproduction (purely sexual species)
12. **One row per species** — keeps the first mean matrix if multiple populations exist

## Inspect what was selected

```{r inspect-db}
# Number of species retained
nrow(db_sexual)

# Species and family per row
as.data.frame(db_sexual) |>
  select(SpeciesAccepted, Family, Order, Country, MatrixDimension) |>
  arrange(Family, SpeciesAccepted) |>
  print(n = Inf)

# Family representation
sort(table(as.character(db_sexual$Family)), decreasing = TRUE)
```

## Save the filtered database to avoid re-downloading

COMPADRE is large (~200 MB download). Save the filtered object once and
reload it in future sessions.

```{r save-db}
# Save after filtering (run once)
saveRDS(db_sexual, "db_sexual_compadre.rds")

# Reload in future sessions instead of running the filter again
# db_sexual <- readRDS("db_sexual_compadre.rds")
```

---

# Step 2: Run the Ne analysis

`Ne_compadre_sexual()` loops over every species in `db_sexual`, computes
the stable stage distribution $D$, then calls `Ne_sexual_Y2000()`.

```{r run-ne}
out <- Ne_compadre_sexual(
  db        = db_sexual,
  Ne_target = 5000,    # Ne conservation threshold; adjust as needed
  Vk_over_k = NULL,    # NULL = Poisson variance (Vk/k = 1); see Section 5
  verbose   = TRUE     # print one line per species
)
```

The output is a named list with two elements:

| Element | Type | Content |
|---|---|---|
| `out$summary` | `data.frame` | One row per species — see column glossary below |
| `out$results` | named `list` | Full `Ne_sexual_Y2000` result object per species |

### Column glossary for `out$summary`

| Column | Description |
|---|---|
| `Species` | Accepted species name |
| `Family` | Plant family |
| `Order` | Taxonomic order |
| `Country` | Country of study |
| `Lat`, `Lon` | Study site coordinates |
| `MatrixDimension` | Number of life-history stages $s$ |
| `lambda` | Population growth rate $\lambda = \max \text{Re}(\text{eig}(A))$ |
| `NeN` | $N_e / N$ — effective to census size ratio |
| `Min_N` | Minimum census $N$ for $N_e \geq N_{e,\text{target}}$ |
| `L` | Generation time (years), computed from $U + F$ |
| `V` | Variance in lifetime reproductive success |
| `u_bar` | Population mean annual survival $\bar{u}$ |
| `u2_bar` | Stage-weighted mean of squared survivals $\overline{u^2}$ |
| `Ne_target` | The $N_e$ threshold used |

---

# Step 3: Inspect and interpret results

## Summary statistics

```{r stats}
summary_df <- out$summary

cat("Species analysed: ", nrow(summary_df), "\n")
cat("Families covered: ", length(unique(summary_df$Family)), "\n")
cat("Ne/N range:       ",
    round(min(summary_df$NeN), 4), "--",
    round(max(summary_df$NeN), 4), "\n")
cat("Median Ne/N:      ", round(median(summary_df$NeN), 4), "\n")
cat("Min_N range:      ", min(summary_df$Min_N), "--",
    max(summary_df$Min_N), "\n")
```

## Full results table

```{r full-table}
print(summary_df, digits = 4)
```

## Species of conservation concern

Species with $N_e/N < 0.10$ fall below Frankham's (1995) empirical median
across wildlife species, indicating unusually low genetic representation per
individual in the census population.

```{r low-ne}
summary_df |>
  filter(NeN < 0.10) |>
  arrange(NeN) |>
  select(Species, Family, NeN, Min_N, L, lambda) |>
  print()
```

## Which species need the largest populations?

Species with low $N_e/N$ require very large census populations to maintain
$N_e \geq 5000$. These are candidates for priority conservation action.

```{r large-minn}
summary_df |>
  arrange(desc(Min_N)) |>
  slice_head(n = 10) |>
  select(Species, Family, NeN, Min_N, L, MatrixDimension) |>
  print()
```

## Accessing a full single-species result

Each element of `out$results` is the complete object returned by
`Ne_sexual_Y2000()`, identical to what you would get running the function
manually for that species.

```{r single-species}
# Print the full result for the first species in the list
first_sp <- names(out$results)[1]
cat("Full Ne result for:", first_sp, "\n\n")
print(out$results[[first_sp]])
```

You can also access any species by name:

```{r single-by-name}
# Replace with any species name in your results
# print(out$results[["Cirsium pitcheri"]])
```

## Family-level summary

```{r family-summary}
fam_summary <- summary_df |>
  group_by(Family) |>
  summarise(
    n_species  = n(),
    median_NeN = round(median(NeN), 4),
    min_NeN    = round(min(NeN),    4),
    max_NeN    = round(max(NeN),    4),
    mean_L     = round(mean(L),     2),
    .groups    = "drop"
  ) |>
  arrange(median_NeN)

print(fam_summary, n = Inf)
```

---

# Step 4: Plots

## Plot 1 — Distribution of $N_e/N$ across all species

```{r plot1, fig.width=7, fig.height=4.5}
ggplot(summary_df, aes(x = NeN)) +
  geom_histogram(bins = 30, fill = "#2166ac", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = 0.10, linetype = "dashed",
             colour = "#d6604d", linewidth = 0.9) +
  annotate("text", x = 0.105, y = Inf,
           label = "Frankham (1995)\nmedian = 0.10",
           hjust = 0, vjust = 1.6, size = 3.2, colour = "#d6604d") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title    = expression("Distribution of " * N[e]/N * " — COMPADRE plant species"),
    subtitle = "Sexually reproducing species | mean matrices | wild, unmanipulated",
    x        = expression(N[e] / N),
    y        = "Number of species",
    caption  = "Yonezawa et al. (2000) Eq. 6, d = 0. D from stable stage distribution."
  ) +
  theme_classic(base_size = 13)
```

**How to read this:** Each bar is a bin of species with similar $N_e/N$. The
red dashed line is the Frankham (1995) empirical median of 0.10 across
wildlife species. Species to the left of the line have lower effective
population sizes relative to their census size — they lose genetic diversity
faster per individual counted.

## Plot 2 — $N_e/N$ by plant family

Only families with ≥ 2 species are shown so that distributions are
meaningful.

```{r plot2, fig.width=7.5, fig.height=6}
fam_min2 <- names(which(table(summary_df$Family) >= 2))
df_fam   <- summary_df |> filter(Family %in% fam_min2)

ggplot(df_fam,
       aes(x = reorder(Family, NeN, FUN = median), y = NeN)) +
  geom_boxplot(fill = "#4dac26", alpha = 0.55,
               outlier.shape = NA, width = 0.55) +
  geom_jitter(width = 0.15, size = 2.0, alpha = 0.75, colour = "#1b7837") +
  geom_hline(yintercept = 0.10, linetype = "dashed",
             colour = "#d6604d", linewidth = 0.8) +
  annotate("text", x = 0.65, y = 0.104,
           label = "Frankham 0.10", hjust = 0,
           size = 3.0, colour = "#d6604d") +
  coord_flip() +
  labs(
    title    = expression("Ne/N by plant family (COMPADRE)"),
    subtitle = "Families with \u2265 2 species",
    x        = NULL,
    y        = expression(N[e] / N),
    caption  = "Each point = one species. Boxes show median and IQR."
  ) +
  theme_classic(base_size = 13)
```

**How to read this:** Each box covers the interquartile range for that
family. Families sorted left to right by median $N_e/N$. Points above the
red dashed line have above-average genetic representation per census individual.

## Plot 3 — $N_e/N$ vs generation time $L$

A longer generation time generally means more opportunity for genetic drift
between reproductive events, but also fewer reproductive events per year,
so the relationship between $L$ and $N_e/N$ is not fixed.

```{r plot3, fig.width=8, fig.height=5.5}
ggplot(summary_df,
       aes(x = L, y = NeN, colour = Family)) +
  geom_smooth(
    mapping     = aes(x = L, y = NeN),
    method      = "lm", se = TRUE,
    colour      = "grey50", fill = "grey85",
    linetype    = "dashed", linewidth = 0.7,
    inherit.aes = FALSE
  ) +
  geom_point(size = 2.5, alpha = 0.85) +
  geom_hline(yintercept = 0.10, linetype = "dotted",
             colour = "#d6604d", linewidth = 0.6) +
  labs(
    title    = expression(N[e]/N * " vs generation time " * italic(L)),
    subtitle = "Each point = one species, coloured by family",
    x        = "Generation time L (years)",
    y        = expression(N[e] / N),
    colour   = "Family",
    caption  = "Red dotted line = Frankham (1995) median."
  ) +
  theme_classic(base_size = 13) +
  theme(
    legend.position  = "right",
    legend.text      = element_text(size = 7),
    legend.key.size  = unit(0.35, "cm"),
    legend.title     = element_text(size = 8)
  )
```

## Plot 4 — Minimum census size needed, by family

This is the most direct conservation output: how many individuals must be
present for the effective population size to reach the threshold of 5000?

```{r plot4, fig.width=7.5, fig.height=6}
ggplot(summary_df,
       aes(x = reorder(Family, Min_N, FUN = median), y = Min_N)) +
  geom_point(colour = "#762a83", size = 2.5, alpha = 0.8) +
  geom_hline(yintercept = 5000, linetype = "dashed",
             colour = "grey40", linewidth = 0.7) +
  annotate("text", x = 0.65, y = 5200,
           label = "Ne target = 5000", hjust = 0, vjust = 0,
           size = 3.0, colour = "grey40") +
  scale_y_log10(labels = scales::comma) +
  coord_flip() +
  labs(
    title    = expression("Minimum census size for " * N[e] >= 5000),
    subtitle = "One point per species, sorted by family median",
    x        = NULL,
    y        = "Minimum N (log scale)",
    caption  = "Min N = Ne_target / (Ne/N)."
  ) +
  theme_classic(base_size = 13)
```

---

# Step 5: Sensitivity to reproductive variance ($V_k/\bar{k}$)

The Poisson default (`Vk_over_k = NULL`, which sets $V_k/\bar{k} = 1$) is
the most conservative assumption when individual-level variance data are
unavailable. If your study species shows over-dispersed reproduction —
common in long-lived plants where only dominant adults reproduce in a given
year — a larger value will reduce $N_e/N$ further.

```{r vk-sensitivity}
# Re-run with over-dispersed reproduction (Vk/k = 2)
out_v2 <- Ne_compadre_sexual(
  db        = db_sexual,
  Ne_target = 5000,
  Vk_over_k = 2,        # variance twice the mean
  verbose   = FALSE
)

# Compare Ne/N under Poisson vs over-dispersed
comparison <- data.frame(
  Species    = out$summary$Species,
  Family     = out$summary$Family,
  NeN_pois   = out$summary$NeN,
  NeN_vk2    = out_v2$summary$NeN,
  NeN_ratio  = round(out_v2$summary$NeN / out$summary$NeN, 3)
)

print(comparison)
```

As a rule of thumb, doubling $V_k/\bar{k}$ from 1 to 2 typically reduces
$N_e/N$ by 20--40%, depending on how much of $V$ is driven by the
reproductive variance term vs the survival variance term.

---

# Step 6: Save results

```{r save-results}
write.csv(out$summary,  "Ne_compadre_sexual_results.csv",  row.names = FALSE)
write.csv(fam_summary,  "Ne_compadre_family_summary.csv",  row.names = FALSE)

# Optional: save the full results list as an R object
saveRDS(out, "Ne_compadre_sexual_out.rds")

cat("Saved:\n")
cat("  Ne_compadre_sexual_results.csv  (one row per species)\n")
cat("  Ne_compadre_family_summary.csv  (family-level statistics)\n")
cat("  Ne_compadre_sexual_out.rds      (full R object)\n")
```

---

# Interpreting $N_e/N$

| $N_e/N$ | Interpretation |
|---|---|
| > 0.25 | High — effective size close to census size |
| 0.10 -- 0.25 | Moderate — near Frankham (1995) empirical median for wildlife |
| 0.01 -- 0.10 | Low — substantial genetic drift; conservation concern |
| < 0.01 | Very low — strong genetic drift; high extinction risk unless $N$ is very large |

Reference thresholds:

- **Frankham (1995):** median empirical $N_e/N \approx 0.10$ across wildlife species
- **Franklin (1980) / Lande (1995) rule:** $N_e \geq 50$ to avoid inbreeding
  depression in the short term; $N_e \geq 500$--$5000$ for long-term evolutionary
  potential

---

# Notes and limitations

**D from stable stage distribution.** The stage frequency vector $D$ is
computed from the dominant right eigenvector of $A = U + F$. This represents
the theoretical equilibrium. If a population is far from equilibrium (e.g.,
recovering from a disturbance, or declining rapidly), the realised $N_e/N$
will differ. Where field census data on stage structure are available,
supply $D$ directly to `Ne_sexual_Y2000()`.

**One matrix per species.** When multiple mean matrices exist for the same
species (different populations or time periods), only the first is retained.
Substantial among-population variance in $N_e/N$ is possible. To explore
this, pass a less-filtered `CompadreDB` object directly to
`Ne_compadre_sexual()`.

**Poisson reproductive variance.** The default $V_k/\bar{k} = 1$ is
conservative. Perennial plants often have highly over-dispersed reproduction
(few individuals reproduce in any year), which would lower $N_e/N$ further.
See Section 5 for a sensitivity analysis.

**Annual matrices only.** `AnnualPeriodicity == 1` is enforced because
NeStage equations assume a 1-year time step. Sub-annual (seasonal) matrices
are excluded.

---

# References

Franklin, I. R. (1980). Evolutionary change in small populations. In
*Conservation Biology: An Evolutionary-Ecological Perspective* (eds
Soulé & Wilcox), pp. 135--149. Sinauer, Sunderland MA.

Frankham, R. (1995). Effective population size/adult population size ratios
in wildlife: a review. *Genetics Research*, 66, 95--107.

Jones, O. R., et al. (2022). Rcompadre and Rage: Two R packages to
facilitate the use of the COMPADRE and COMADRE databases.
*Methods in Ecology and Evolution*, 13, 1164--1169.
doi:[10.1111/2041-210X.13792](https://doi.org/10.1111/2041-210X.13792)

Lande, R., & Barrowclough, G. F. (1987). Effective population size,
genetic variation, and their use in population management. In
*Viable Populations for Conservation* (ed. Soulé), pp. 87--123.
Cambridge University Press.

Salguero-Gómez, R., et al. (2015). The COMPADRE Plant Matrix Database:
an open online repository for plant demography. *Journal of Ecology*,
103, 202--218.

Yonezawa, K., Ishii, T., & Nagamine, T. (2000). The effective size of
stage-structured populations with selfing or clonal reproduction.
*Evolution*, 54, 1707--1716.
doi:[10.1111/j.0014-3820.2000.tb01243.x](https://doi.org/10.1111/j.0014-3820.2000.tb01243.x)

---

```{r session-info}
sessionInfo()
```
